Uncertainty in social learning evolution

### 2022-05-22 ms.tex
first full methods, analysis, +, draft

\begin{figure}
  \caption{Agents are initialized as randomly social learners or not, with their
  payoff observations all initialized to zero (A). Then agents begin selecting
and performing behaviors and accumulating payoffs (1), which goes on for $L$
timesteps (B). After $L$ time steps (2), agents are selected to reproduce,
social learner children learn from a member of their parent's generation, and
the previous generation dies off (C). Stop the simulation if children are all
social or asocial learners, otherwise repeat another generation and evolution (3).}
  \label{fig:schematic}
  \centering
    \includegraphics[width=\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

In between generations agents first reproduce via asexual haploid reproduction; 
then new child agents learn from the
previous generation if they are social learners; and finally all agents from the
previous generation then die off. $N$ reproducers are selected with replacement, 
biased by performance:
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
Child agents inherit their parents social learning trait $s_i$ without mutation.
Child agents with $s_i = 1$ then learn from one teacher from their parent's
generation, including possibly their parent. Child social learners select
their teacher first by selecting $N_T$ prospective teachers from the 
population completely at random, then selecting the one with highest
net payoffs $\pi_i$ for as their teacher, with ties broken randomly 
($N_T = 5$ in the main text;
\mt{Other values will be tested in the supplement}). While it is somewhat
arbitrary to first subset $N_T$, this is a conservative choice that
represnts the fact that access to the entire population is not generally
guaranteed. Social learners adopt their chosen teacher's ledger of 
expected payoffs $\ledger$, and set
all $c_{ib} = 1$ so expected payoffs remain within $[0, 1]$. 
Asocial learners ($s_i = 0$) are initialized with all $c_{ib} = 0$ and
$\ledger = 0$. The newly initialized population then begins behavior selection
and payoff accumulation for another $L$ time steps and the evolutionary process
continues until $\sum_i s_i = 1$ or $\sum_i s_i = 0$ \mt{I believe
all trials will eventually end like this, so in the next set of model runs I will
not allow early termination before one of these conditions are met}.

@@

We observed several patterns in the suppression of social learning as environmental
variability, $u$, increased across different uncertainty contexts defined by the
payoff ambiguity, selection set size, and effective lifespan
(Figure~\ref{fig:mainResults}). When payoff ambiguity and selection set size
were both small ($\pilow = 0.1$ and $B=2$) social learning only evolved when
$u$ is small since individual learning is relatively efficient in this case
(Figure~\ref{fig:mainResults} upper left). As $\pilow$ increased the suppression of
$\meansl$ flattened over $u$, which was more pronounced for smaller $L$ (rows of
Figure~\ref{fig:mainResults}). 
When $L$ is smaller, agents have fewer opportunities to learn through trial and
error, and so it is more difficult for them to discern which strategy, social
or asocial, is optimal. Increased selection set size led to both a shift in 
the critical value of $u$ when $\meansl$ begins to decrease, and flattening of
this social learning suppression. The start of $\meansl$ suppression shifts
to higher $u$ since misleading social information is less consequential due to
(1) larger $B$ being a more difficult problem for trial-and-error learning and
(2) less probability mass of a misidentified optimal behavior relative to
smaller $B$, meaning agents are more likely to explore behaviors identified
as non-optimal (columns of Figure~\ref{fig:mainResults}). 
Increased $B$ led to more pronounced flattening of social learning
suppression when $L$ was small, apparently again because of the limited data
available to evolution and more difficult problem caused more drift.

@@

\paragraph{Analysis 2: how payoffs and generations to convergence support our
explanations}

\begin{figure}
  \caption{Mean payoff normalized for lifespan (y-axes, solid lines with circles)
    generally track either expected all-social-learners payoff ($\meansoc$, dash-dot
    line with diamonds) or expected all-asocial payoff ($\meanasoc$, long dash
    horizontal lines) as $u$ increases (x-axes). When $\meansl$ goes from 1 to
  0 is when social learning payoffs approach individual learning payoffs due to
increased environmental variability.} 
  \label{fig:netPayoffs}
\centering
    \includegraphics[width=0.75\textwidth]{Figures/meanNetPayoffs.pdf}
\end{figure}

>>>>

\begin{figure}
  \caption{Agents are initialized as randomly social learners or not, with their
  payoff observations all initialized to zero (A). Then agents begin selecting
and performing behaviors and accumulating payoffs, which goes on for $L$
timesteps (B). After $L$ time steps, agents are selected to reproduce,
social learner children learn from a member of their parent's generation, and
the previous generation dies off (C). The simulation stops if children are all
social or asocial learners (i.e.\ the system reaches fixation), 
otherwise repeat another generation and evolution (3).}
  \label{fig:schematic}
  \centering
    \includegraphics[width=\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

In between generations agents first reproduce via asexual haploid reproduction; 
then new child agents learn from the
previous generation if they are social learners; and finally all agents from the
previous generation then die off. $N$ reproducers are selected with replacement, 
biased by performance:
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
Child agents inherit their parent's social learning trait $s_i$ without mutation.
Child agents with $s_i = 1$ then learn from one teacher from their parent's
generation, including possibly their parent. Child social learners select
their teacher first by selecting $N_T$ prospective teachers from the 
population completely at random, then selecting the one with highest
net payoffs $\pi_i$ for as their teacher, with ties broken randomly 
($N_T = 5$ in the main text;
\mt{Other values will be tested in the supplement}). While it is somewhat
arbitrary to first subset $N_T$, this is a conservative choice that
represnts the fact that access to the entire population is not generally
guaranteed. Social learners adopt their chosen teacher's ledger of 
expected payoffs $\ledger$, and set
all $c_{ib} = 1$ so expected payoffs remain within $[0, 1]$. 
Asocial learners ($s_i = 0$) are initialized with all $c_{ib} = 0$ and
$\ledger = 0$. The newly initialized population then begins behavior selection
and payoff accumulation for another $L$ time steps and the evolutionary process
continues until $\sum_i s_i = 1$ or $\sum_i s_i = 0$ \mt{I believe
all trials will eventually end like this, so in the next set of model runs I will
not allow early termination before one of these conditions are met}.

@@

We observed several patterns in the suppression of social learning as environmental
variability, $u$, increased across different uncertainty contexts defined by the
payoff ambiguity, selection set size, and effective lifespan
(Figure~\ref{fig:mainResults}). When payoff ambiguity and selection set size
were both small ($\pilow = 0.1$ and $B=2$) social learning only evolved when
$u$ is small since individual learning is relatively efficient in this case
(Figure~\ref{fig:mainResults} upper left). As $\pilow$ increased the suppression of
$\meansl$ flattened over $u$, which was more pronounced for smaller $L$ (rows of
Figure~\ref{fig:mainResults}). 
When $L$ is smaller, agents have fewer opportunities to learn through trial and
error, and so it is more difficult for them to discern which strategy, social
or asocial, is optimal. Increased selection set size led to both a shift in 
the critical value of $u$ when $\meansl$ begins to decrease, and flattening of
this social learning suppression. The start of $\meansl$ suppression shifts
to higher $u$ since misleading social information is less consequential due to
(1) larger $B$ being a more difficult problem for trial-and-error learning and
(2) less probability mass of a misidentified optimal behavior relative to
smaller $B$, meaning agents are more likely to explore behaviors identified
as non-optimal (columns of Figure~\ref{fig:mainResults}). 
Increased $B$ led to more pronounced flattening of social learning
suppression when $L$ was small, apparently again because of the limited data
available to evolution and more difficult problem caused more drift and individual
outcomes were more path dependent.

@@

We observed that the evolution of social learning was correlated with
population-level payoffs for social versus asocial learning. Furthermore, we
found that mean payoffs realized by our simulated agents track either the
expected social learning payoffs, $\meansoc$, or the expected asocial learning
payoffs, $\meanasoc$. $\meanasoc$ and $\meansoc$ were calculated
across uncertainty contexts by initializing the model to be all social or all
individual learners then running the model for 100 generations for social learning
and across 1000 one-generation trials for individual learning. 
Observed mean payoffs across agents and trials, $\meanpi$, are initially higher than
expected asocial learning, $\meanasoc$, and comparable to expected social learning,
$\meansoc$ (Figure~\ref{fig:payoffs}). But as social learning becomes less
beneficial due to increased environmental variability, $\meansoc$ and $\meanpi$
approach $\meanasoc$. It appears agents tend to be conservative as expected
social and asocial learning payoffs approach each other---the transition to 
$\meansl = 0$ often completes in cases where social learning is theoretically
better. When $L=1$, $\meanpi$ can be greater than $\meansoc$
because asocial learners at the beginning of the simulation help to not pass
on misleading information, so that the cultural information passed on by 
eventual all-social learner populations have more noise that will further reduce the
weighting of the previous generation's observed optimal payoff if it changes
between generations enabling agents to more efficiently explore alternatives
before focusing on a misidentified optimal behavior \mt{I think we can do a
real calculation of this if it would be valuable}.

\begin{figure}
  \caption{Mean payoff normalized for lifespan (y-axes, solid lines with circles)
    generally track either expected all-social-learners payoff ($\meansoc$, dash-dot
    line with diamonds) or expected all-asocial payoff ($\meanasoc$, long dash
    horizontal lines) as $u$ increases (x-axes). When $\meansl$ goes from 1 to
  0 is when social learning payoffs approach individual learning payoffs due to
increased environmental variability.} 
  \label{fig:payoffs}
\centering
    \includegraphics[width=0.75\textwidth]{Figures/meanNetPayoffs.pdf}
\end{figure}

We also observed that the number of generations it took to reach fixation increased
when expected social learning payoffs approached expected asocial payoff, which is also
when $\meansl$ decreased from 1 to 0 (Figure~\ref{fig:steps}). Recall a fixated
state is when $\sum_i s_i = 0 \text{ or } 0$.  This supports our explanation that
the supression of social learning begins when individual simulation outcomes are
more path dependent and susceptible to end in a non-optimal state since simulations
in these conditions also take longer to reach fixation.  For example, when
$\pilow=0.1$ and $B=4$, we see a sharp increase in $\meanT$ when social learning
suppression begins and $L=8$; when $L=1$, social learning suppression is flatter
than for $L=8$, and  correspondingly its $\meanT$ is not as strongly peaked, but is
elevated across the $u$ (Figure~\ref{fig:steps} upper left).  In some cases $L=1$
reaches fixation faster than $L=8$ or $L=20$; larger $L$ had larger peak $\meanT/L$
than smaller $L$, for example, both for the cases shown in Figure~\ref{fig:steps}
and across all settings tested but not shown. However, we also see situations where
$L=1$ and fixation takes longer than $L=8$.  This is especially pronounced for
$\pilow = 0.45$ where social learning suppression is much flatter for $L=1$
(Figure~\ref{fig:steps}, bottom row). In general, then, a sharper peak in $\meanT$
indicates a sharper suppression of social learning over $u$ and a flatter peak
elevated over more values of $u$ indicates greater path dependence and drift.
